---
title: "COVID19 Tends in Eastern and Southern Africa region"
author: "hrayr"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen=999)
require("sfsmisc")
library(knitr)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(reshape2)
library(sf)
library(tmap)
library(rgdal)
library(tidyverse)
### get population data
library(wpp2019)
data(pop)
esapop<- pop[pop$name %in% c("Angola"
                                     ,"Botswana"
                                     ,"Burundi"
                                     ,"Comoros"
                                     ,"Eritrea"
                                     ,"Eswatini"
                                     ,"Ethiopia"
                                     ,"Kenya"
                                     ,"Lesotho"
                                     ,"Madagascar"
                                     ,"Malawi"
                                     ,"Mozambique"
                                     ,"Namibia"
                                     ,"Rwanda"
                                     ,"Somalia"
                                     ,"South Africa"
                                     ,"South Sudan"
                                     ,"United Republic of Tanzania"
                                     ,"Uganda"
                                     ,"Zambia"
                                     ,"Zimbabwe"),]

esapop<- esapop[,names(esapop) %in% c("name","2020")]
names(esapop)<- c("country","Population_2020 (000)")
esapop$country<-as.character(esapop$country)

esapop[esapop=="United Republic of Tanzania"]<- "Tanzania"
###get esa shape
esamap<-read_sf("C:/USERS/hwannis/OneDrive - UNICEF/MENARO/91. ESARO/GIS/ESA_admin1_region.shp")
esamap<- esamap[esamap$COUNTRY %in% esapop$country,]

#esamap[esamap$COUNTRY=="Iran",1]<-c("Iran (Islamic Republic of)")
#esamap[esamap$COUNTRY=="Syria",1]<-c("Syrian Arab Republic")
esamap<-merge(esamap,esapop,by.x="COUNTRY",by.y="country")

covid19_new<-read.csv("https://covid.ourworldindata.org/data/owid-covid-data.csv", header = TRUE, as.is = TRUE) 

covid19_new %>% select("location","date","total_cases") %>% filter((location!= "")) %>% pivot_wider(names_from = location,values_from = total_cases,values_fill = list(total_cases=0)) %>% arrange(date)->covid19_db

#remove last day/today in database given it is not complete at times
covid19_db<-head(covid19_db,-1)

#remove NA in between dates
testdb<-covid19_db
for(x in 1:ncol(testdb)){
    for (i in 2:nrow(testdb)){
      if (is.na(testdb[[i,x]]) & !is.na(testdb[[i-1,x]])) {
      testdb[[i,x]]=testdb[[i-1,x]]
    }   else if (is.na(testdb[[i,x]])) {
      testdb[[i,x]]=0
    } else {}
    
  }
  if (is.na(testdb[[1,x]])) { testdb[[1,x]]=0}
  }

covid19_db<-testdb

#covid19_deaths<-read.csv("https://covid.ourworldindata.org/data/ecdc/total_deaths.csv", header=TRUE, as.is=TRUE)

covid19_new %>% select("location","date","total_deaths") %>% filter((location!= "")) %>% pivot_wider(names_from = location,values_from = total_deaths,values_fill = list(total_deaths=0))%>% arrange(date)->covid19_deaths

#remove NA in between dates
testdb<-covid19_deaths
for(x in 1:ncol(testdb)){
  for (i in 2:nrow(testdb)){
      if (is.na(testdb[[i,x]]) & !is.na(testdb[[i-1,x]])) {
      testdb[[i,x]]=testdb[[i-1,x]]
    }   else if (is.na(testdb[[i,x]])) {
      testdb[[i,x]]=0
    } else {}
    
  }
  if (is.na(testdb[[1,x]])) { testdb[[1,x]]=0}
  }

covid19_deaths<-testdb
#remove last day/today in database given it is not complete at times
covid19_deaths<-head(covid19_deaths,-1)
#names(covid19_db)[names(covid19_db)=="Palestine"]<-c("State of Palestine")
#names(covid19_db)[names(covid19_db)=="Iran"]<-c("Iran (Islamic Republic of)")
#names(covid19_db)[names(covid19_db)=="United.Arab.Emirates"]<-c("United Arab Emirates")
#names(covid19_db)[names(covid19_db)=="Syria"]<-c("Syrian Arab Republic")
#names(covid19_db)[names(covid19_db)=="Saudi.Arabia"]<-c("Saudi Arabia")


covid19_esa<-covid19_db[names(covid19_db) %in% c("date"
                                     ,"Angola"
                                     ,"Botswana"
                                     ,"Burundi"
                                     ,"Comoros"
                                     ,"Eritrea"
                                     ,"Eswatini"
                                     ,"Ethiopia"
                                     ,"Kenya"
                                     ,"Lesotho"
                                     ,"Madagascar"
                                     ,"Malawi"
                                     ,"Mozambique"
                                     ,"Namibia"
                                     ,"Rwanda"
                                     ,"Somalia"
                                     ,"South Africa"
                                     ,"South Sudan"
                                     ,"Tanzania"
                                     ,"Uganda"
                                     ,"Zambia"
                                     ,"Zimbabwe")]

#names(covid19_deaths)[names(covid19_deaths)=="Palestine"]<-c("State of Palestine")
#names(covid19_deaths)[names(covid19_deaths)=="Iran"]<-c("Iran (Islamic Republic of)")
#(covid19_deaths)[names(covid19_deaths)=="United.Arab.Emirates"]<-c("United Arab Emirates")
#names(covid19_deaths)[names(covid19_deaths)=="Syria"]<-c("Syrian Arab Republic")
#names(covid19_deaths)[names(covid19_deaths)=="Saudi.Arabia"]<-c("Saudi Arabia")


covid19_esad<-covid19_deaths[names(covid19_deaths) %in% c("date"
                                     ,"Angola"
                                     ,"Botswana"
                                     ,"Burundi"
                                     ,"Comoros"
                                     ,"Eritrea"
                                     ,"Eswatini"
                                     ,"Ethiopia"
                                     ,"Kenya"
                                     ,"Lesotho"
                                     ,"Madagascar"
                                     ,"Malawi"
                                     ,"Mozambique"
                                     ,"Namibia"
                                     ,"Rwanda"
                                     ,"Somalia"
                                     ,"South Africa"
                                     ,"South Sudan"
                                     ,"Tanzania"
                                     ,"Uganda"
                                     ,"Zambia"
                                     ,"Zimbabwe")]

##france trend data
covid19_fr<-covid19_db[names(covid19_db) %in% c("France")]

##italy trend data
covid19_italy<-covid19_db[names(covid19_db) %in% c("Italy")]

##spain trend data
covid19_spain<-covid19_db[names(covid19_db) %in% c("Spain")]


##china trend data
covid19_china<-covid19_db[names(covid19_db) %in% c("China")]

##USA trend data
covid19_usa<-covid19_db[names(covid19_db) %in% c("United States")]

##india trend data
covid19_india<-covid19_db[names(covid19_db) %in% c("India")]

##world trend
covid19_wld<-covid19_db[names(covid19_db) %in% c("World")]


#preparing for ESA aggregate
covid19esar<- covid19_esa
covid19esar$ESA<- rowSums(covid19_esa[,-1])
covid19esar<- covid19esar[names(covid19esar) %in% c("date","ESA")]

#bind into datasets mena countries and comparatives
covid19_esa_comparative<-cbind(covid19_esa,covid19_china,covid19_fr,covid19_italy,covid19_spain,covid19_usa, covid19_india)
#bind mena region and comparatives
covid19esar<-cbind(covid19esar,covid19_china,covid19_fr,covid19_italy,covid19_spain, covid19_usa, covid19_india)

#transfor date var into as.Date
covid19<-covid19_esa_comparative
covid19$date<-as.Date(covid19$date)
esac19<-covid19esar
esac19$date<-as.Date(esac19$date)

vt<-covid19$date
vtm<-esac19$date


#for ggploting
esaregional<-melt(esac19,id="date")

covid19$date<-NULL
esac19$date<-NULL

#create incidendce dataframe
c19esa_inci<- covid19-mutate_all(covid19,lag)
c19esa_inci<-cbind(date=vt[-1],c19esa_inci[-1,])
#group by weeks
#c19esa_inci$datew<-format(c19esa_inci$date,"%W")
#c19esa_inci$datey<- format(c19esa_inci$date,"%y")

#date format to year and week-count and summarize counts by week
c19esa_inci$date<- format(c19esa_inci$date,format="%Y-%U")
c19esa_inci_weeks<- c19esa_inci %>% group_by(date) %>% summarise_all(sum)

ndaysinweek<-c19esa_inci %>% group_by(date) %>% dplyr::summarise(daysinweek=n())  

if (ndaysinweek$daysinweek[nrow(ndaysinweek)]<7)  
                          {c19esa_inci_weeks<-c19esa_inci_weeks[1:(nrow(c19esa_inci_weeks)-1),]} else    
                          {c19esa_inci_weeks<-c19esa_inci_weeks}

vtweeks<-seq(1,length(c19esa_inci_weeks$date))

#covid19[,1:ncol(covid19)]<-sapply(covid19[, 1:ncol(covid19)], as.numeric)
#menac19<-as.data.frame(apply(menac19,2, as.numeric))

covid19esa<-covid19[,!names(covid19) %in% c("China","France","Italy","Spain","United States", "India")]
covid19china<-(covid19$China)
covid19italy<-(covid19[,names(covid19) %in% c("Italy")])
covid19spain<-(covid19$Spain)
covid19usa<-covid19$`United States`
row.names(covid19esa)<-NULL

esa_total_deaths<-covid19_esad[nrow(covid19_esad),]
esa_total_deaths$date<-NULL

#preparing data for tabulating
tabing<-as.data.frame(t(covid19esa[nrow(covid19esa),]))
tabing$proportion<-round(tabing[,1]/sum(tabing[,1])*100,2)

tabing<-merge(tabing,esapop,by.x="row.names",by.y="country")
names(tabing)<-c("Country","Tot infected","Percent","Pop (000)")
tabing$TotCases_per1000<-tabing$`Tot infected`/tabing$`Pop (000)`

tabing<-tabing[order(tabing$`Tot infected`,decreasing = TRUE),]

barcharting<-data.frame(names(covid19esa),(t(covid19esa[nrow(covid19esa),])),(t(esa_total_deaths)))
row.names(barcharting)<-NULL                     
names(barcharting)<-c("Country","Confirmed", "Deaths")
cfr<-barcharting


chartingwithpop<-merge(barcharting,esapop,by.x = "Country",by.y="country")

##add vars for mapping
esamap<-merge(esamap,tabing,by.x="COUNTRY",by.y="Country")

```

## ESA Country C19 cases - Data as of `r format(max(vt), '%d %B, %Y')`

The following report is based on data compiled by worldometer from John Hopkins University. The total cumulative number of confirmed COVID19 cases in ESA region is `r max(esac19$ESA)`. The total number of deaths are `r sum(barcharting$Deaths)`. The ESA average infected is `r round(sum(barcharting$Confirmed)/sum(esapop[,2]),2)`per thousand people.

`r kable(tabing, digits=1,row.names=0, caption="Number and Proportion of Infecteds in ESA by Country", longtable = TRUE, format="pandoc")`

```{r ggploting, echo=FALSE,fig.height = 5, fig.width = 7, fig.align = "center"}


dfm <- melt(barcharting[,c("Country","Confirmed","Deaths")],id.vars = 1)

plot1<- ggplot()+geom_col(aes(x=reorder(Country,-value),y=value, fill=variable),data=dfm, position = "dodge")+ theme_economist(base_family="ITC Officina Sans") + scale_colour_economist()+ theme(plot.title=element_text(size=12),legend.title = element_blank(),legend.text = element_text(size=8),legend.key.size = unit(.8,units = "lines"),legend.position = c(.8,.81),axis.title.y = element_blank())+labs(title = paste("ESA COVID19 Confirmed Cases and Deaths - ",format(max(vt), '%d %B, %Y')),y="Number of People")


plot1+coord_flip()

```

The largest share of confirmed cases in ESA are in `r barcharting$Country[barcharting$Confirmed==max(barcharting$Confirmed)]` as of `r format(max(vt), '%d %B, %Y')`. The highest number of recorded deaths so far has been in `r barcharting$Country[barcharting$Deaths==max(barcharting$Deaths)]` with `r max(barcharting$Deaths)`.

```{r mapping,echo=FALSE,fig.height = 10, fig.width = 7, fig.align = "center"}


tm_shape(esamap)+tm_polygons("TotCases_per1000")+tm_text("COUNTRY",size ="AREA",root = 0,auto.placement = TRUE)+tm_compass()+tm_scale_bar(position=c("right", "bottom"))+tm_layout(main.title = "Total infected per 1000 population by country") +tmap_options(check.and.fix = TRUE)

```



```{r casesdeaths, echo=FALSE, fig.height = 5, fig.width = 7, fig.align = "center"}

cfr$proportionDead<-ifelse(cfr$Confirmed>0,(cfr$Deaths/cfr$Confirmed)*100,0)

cfrplot<-ggplot()+ geom_col(aes(x=reorder(Country,-proportionDead),y=proportionDead),data=cfr)+theme_economist(base_family="ITC Officina Sans") + scale_colour_economist()+theme(plot.title=element_text(size=12),axis.text.x =element_text (size=8))+labs(title = paste("ESA COVID19 Case Fatality Rate (%) - ",format(max(vt), '%d %B, %Y')),y="Percentage of Confirmed cases who died", x="")

cfrplot+coord_flip()
```

```{r ggplotingdaily, echo=FALSE,fig.height = 5, fig.width = 7, fig.align = "center"}


dailyincidence<-as.data.frame(t(c19esa_inci[nrow(c19esa_inci),2:21]))
dailyincidence$Country<-rownames(dailyincidence)
names(dailyincidence)<-c("Incidence","Country")

plotdaily<- ggplot()+geom_col(aes(x=reorder(Country,-Incidence),y=Incidence),data=dailyincidence, position = "dodge")+ theme_economist(base_family="ITC Officina Sans") + scale_colour_economist()+ theme(plot.title=element_text(size=12),legend.title = element_blank(),legend.text = element_text(size=8),legend.key.size = unit(.8,units = "lines"),legend.position = c(.8,.81),axis.title.y = element_blank())+labs(title = paste("ESA new cases on - ",format(max(vt), '%d %B, %Y')),y="Number of cases")

plotdaily+coord_flip()

```

```{r ggplotingweekly, echo=FALSE,warning=FALSE,fig.height = 5, fig.width = 8, fig.align = "center"}

# organize the last two weeks incident counts into a df
weeklyincidence<-as.data.frame(t(c19esa_inci_weeks[(nrow(c19esa_inci_weeks)-1):nrow(c19esa_inci_weeks),2:21]))
#get country names into column
weeklyincidence$Country<-rownames(dailyincidence)
# calculate week on week increase rate for values with at least 10 in previous week to avoid large ratios
weeklyincidence$ratio<-ifelse(weeklyincidence$V1>10,((weeklyincidence$V2/weeklyincidence$V1)-1)*100,NA)
# categorize negative and positive values for charting
weeklyincidence$type<-ifelse(weeklyincidence$ratio<0,"n","y")

plotweekly<- ggplot()+geom_col(aes(x=reorder(Country,-ratio),y=ratio,fill=type),data=weeklyincidence, position = "dodge")+ theme_economist(base_family="ITC Officina Sans") + scale_colour_economist()+ theme(plot.title=element_text(size=12),legend.title = element_blank(),legend.text = element_blank(),legend.position = "none",axis.title.y= element_blank(),axis.ticks = element_line(size = 1))+labs(title = paste("ESA Countries incidence -","Percent change from week to week"),y="Percent")


plotweekly+coord_flip()

```

```{r positivityr, echo=FALSE,fig.height = 5, fig.width = 7, fig.align = "center"}

positivity_dfm<-covid19_new %>% select("location","date", "positive_rate", "total_deaths_per_million") %>% filter(location %in% c("Angola"
                    ,"Botswana"
                    ,"Burundi"
                    ,"Comoros"
                    ,"Eritrea"
                    ,"Eswatini"
                    ,"Ethiopia"
                    ,"Kenya"
                    ,"Lesotho"
                    ,"Madagascar"
                    ,"Malawi"
                    ,"Mozambique"
                    ,"Namibia"
                    ,"Rwanda"
                    ,"Somalia"
                    ,"South Africa"
                    ,"South Sudan"
                    ,"Tanzania"
                    ,"Uganda"
                    ,"Zambia"
                    ,"Zimbabwe"))

for (countryX in c("Angola"
                    ,"Botswana"
                    ,"Burundi"
                    ,"Comoros"
                    ,"Eritrea"
                    ,"Eswatini"
                    ,"Ethiopia"
                    ,"Kenya"
                    ,"Lesotho"
                    ,"Madagascar"
                    ,"Malawi"
                    ,"Mozambique"
                    ,"Namibia"
                    ,"Rwanda"
                    ,"Somalia"
                    ,"South Africa"
                    ,"South Sudan"
                    ,"Tanzania"
                    ,"Uganda"
                    ,"Zambia"
                    ,"Zimbabwe")) {
  
  data_plotting_poistivity<- positivity_dfm %>% filter(location==countryX) %>% arrange("date") %>% filter(!is.na(positive_rate))
  
  plot1<- ggplot(data_plotting_poistivity, aes(x= as.Date(date), y=positive_rate))+geom_line(colour = "red")+ theme_economist(base_family="ITC Officina Sans") + scale_colour_economist()+ theme(plot.title=element_text(size=12),legend.title = element_blank(),legend.text = element_text(size=8),legend.key.size = unit(.8,units = "lines"),legend.position = c(.8,.81),axis.title.y = element_blank())+labs(title = paste("ESA COVID19 Positivity rate - ",(max(positivity_dfm$date)), countryX),y="percent")



plot1
  
  
                    }



```

### ESA regional trend - Comparative with other countries

The trend is cumulative confirmed from `r min(vt)` to `r max(vt)`. As comparative, ESA regional aggregate is plotted with trends of other countries which have had significant case loads so far - `r names(esac19)[!names(esac19) %in% c("ESA")]`.

```{r COVIDesa, echo=FALSE,fig.height = 6, fig.width = 7.5, fig.align = "center"}

options(scipen=999)

##plot all countries
matplot(x = (vtm), y = esac19, type = "l",
        xlab = "Time", ylab = "Cumulative Infected", main = "ESA regional trend - comaprative with other countries",
        lwd = 2, lty = 1, bty = "l", col = 1:ncol(esac19), xaxt='n',ylim = c(0,max(esac19)))
axis.Date(side =1 ,at = vt,labels = format(vt, "%b-%y"))

#matlines(x=vt,y=covid19china,type="l",col =ncol(covid19esa)+1 )

## Add legend
legend(x="left",legend=c(names(esac19)), pch = 1, col = 1:ncol(esac19), bty = "n",horiz = 0,cex = .8)

```

```{r COVID, echo=FALSE,include=FALSE,fig.height = 6, fig.width = 7.5, fig.align = "center"}

options(scipen=999)

##plot all countries
matplot(x = (vt), y = log10(covid19esa), type = "l",
        xlab = "Time", ylab = "Cumulative Infected", main = "ESA Countries",
        lwd = 2, lty = 1, bty = "l", col = 2:ncol(covid19esa), xaxt='n',ylim = c(0,max(log10(covid19esa))))

axis.Date(side =1 ,at = vt,labels = format(vt, "%b-%y"))
#matlines(x=vt,y=covid19china,type="l",col =ncol(covid19esa)+1 )

## Add legend
legend(x="left",legend=c(names(covid19esa)), pch = 1, col = 2:ncol(covid19esa), bty = "n",horiz = 0,cex = .70)

```

## Country comparative trends

Each ESA country (red line) is plotted with background countries (grey lines) to give an idea of the current stage of the epidemic in each country as compared to the rest. The second chart is the weekly incidence trend of each country.

```{r countryprofile, echo=FALSE, fig.height=7, fig.width=9, warning=FALSE}
days_since<-100
totaldb<-list()
countrynames<-c()
for(j in seq(1,ncol(covid19))) {

 cntryseries<-covid19[covid19[[j]]>=days_since,j]

 if (length(cntryseries)<nrow(covid19) & length(cntryseries)>0){
     cntryseries[(length(cntryseries)+1):(nrow(covid19))]<-NA

     assign(paste0(names(covid19)[j]),cntryseries)
     
     totaldb<-cbind(totaldb,assign(paste0(names(covid19)[j]),cntryseries))
     countrynames<-c(countrynames,names(covid19)[j])
 } else if (length(cntryseries)==nrow(covid19))  {
   totaldb<-cbind(totaldb,assign(paste0(names(covid19)[j]),cntryseries))
   countrynames<-c(countrynames,names(covid19)[j])
 }
 totaldb<-as.data.frame(totaldb)

 names(totaldb)<-countrynames
}

 totaldb<-as.data.frame(sapply(totaldb, as.numeric))


options(scipen=999)

vtx<-seq(0,nrow(totaldb)-1)

graphingesaweeks<-c19esa_inci_weeks

graphingesaweeks$date<-NULL

for (y in names(totaldb)){
  mult.fig(2,main=paste("Country Trend - ",y),cex.main = 2)
  
  matplot(x = vtx, y = log10((totaldb)), type = "l", xlab = paste("Days since", days_since, "case"), ylab = "Log scale", main = paste("Cumulative number of confirmed cases per day since",days_since,"cases reported"),lwd = 2, lty = 3, bty = "l", col = "grey", xaxt='n', cex.main=1.4, cex.lab=.9)
  
matlines(x=(vtx), y=log10(totaldb[[paste(y)]]), col="red",lwd = 2, lty = 1, bty = "l",xaxt='n', ylim=c(0,max(((log10(totaldb[[paste(y)]]))),na.rm = TRUE)))
  #par(cex.main=.5)
  axis(side =1, cex.axis=1)
  
  cat("The date when reported case in",y,"went above 100 cases was on",format((as.Date(min(covid19_db$date[covid19_db[[paste(y)]]>100]))),'%d %B, %Y'),"\n")
  
   matplot(x = vtweeks, y = ((graphingesaweeks)), type = "l", xlab = paste("Weeks since begining of Year"), ylab = "People/WeeK", main = "Weekly Incidence",lwd = 2, lty = 3, bty = "l", col = "grey", xaxt='n', cex.main=1.4, cex.lab=.9,ylim=c(0,max((graphingesaweeks[[paste(y)]]))))
  

matlines(x=(vtweeks), y=(graphingesaweeks[[paste(y)]]), col="red",lwd = 2, lty = 1, bty = "l")
  #par(cex.main=.5)
  axis(side =1, cex.axis=1)
  
  
  }

```

```{r casesnew, echo=FALSE, include=FALSE, fig.height = 18, fig.width = 9, fig.align = "center"}
graphingesaweeks<-c19esa_inci_weeks

graphingesaweeks$date<-NULL

#mult.fig(mfrow=c(ncol(graphingesaweeks)/3,3),main=paste("esa Countries - Total weekly Incidence"))

layout(matrix(seq(1,3*ceiling(ncol(graphingesaweeks)/3)),ncol=3),widths = lcm(rep.int(7,3)),height =lcm (rep.int(4.5,round(ncol(graphingesaweeks)/3))), respect = FALSE)

for (y in names(graphingesaweeks)){
  
  matplot(x = vtweeks, y = ((graphingesaweeks)), type = "l", xlab = paste("Weeks"), ylab = "People/WeeK", main = y,lwd = 2, lty = 3, bty = "l", col = "grey", xaxt='n', cex.main=1.2, cex.lab=.9,ylim=c(0,max((graphingesaweeks[[paste(y)]]))))
  
  

matlines(x=(vtweeks), y=(graphingesaweeks[[paste(y)]]), col="red",lwd = 2, lty = 1, bty = "l")
  #par(cex.main=.5)
  axis(side =1, cex.axis=1)
  
}

mtext(text=paste("ESA Countries - Total weekly Incidence"),side = 3, line = 0, outer = TRUE)
```

```{r COVIDesagg,include=FALSE, echo=FALSE,fig.height = 5, fig.width = 9, fig.align = "center"}

options(scipen=999)

##plot all countries
plotesa<- ggplot() + theme_economist(base_family="ITC Officina Sans") + scale_colour_economist()+geom_line(aes(x=date,y=value,colour=variable),data=esaregional)+ theme(plot.title=element_text(size=12),legend.title = element_blank(),legend.text = element_text(size=9),legend.key.size = unit(.9,units = "lines"),legend.position = c(.8,.81))+labs(title = paste("esa region and Comparatives"),y="Number of People")

plotesa

```

```{r COVIDVacc, echo=FALSE, include= TRUE, fig.height = 6, fig.width = 8, fig.align = "center"}
options(scipen=999)

##plot vaccination rates

covid19_new %>% filter(location %in%  c("Angola"
                                     ,"Botswana"
                                     ,"Burundi"
                                     ,"Comoros"
                                     ,"Eritrea"
                                     ,"Eswatini"
                                     ,"Ethiopia"
                                     ,"Kenya"
                                     ,"Lesotho"
                                     ,"Madagascar"
                                     ,"Malawi"
                                     ,"Mozambique"
                                     ,"Namibia"
                                     ,"Rwanda"
                                     ,"Somalia"
                                     ,"South Africa"
                                     ,"South Sudan"
                                     ,"Tanzania"
                                     ,"Uganda"
                                     ,"Zambia"
                                     ,"Zimbabwe","France", "United States","United Kingdom","World"),  
                       !is.na(new_deaths_smoothed_per_million))%>% select(location,date,new_deaths_smoothed_per_million,people_fully_vaccinated_per_hundred) -> vaccinationDf


for (x in unique(vaccinationDf$location)) {
#scale sec axis to first by factor max(2)/max(1)  
vaccinationDf %>% filter(location==x) %>% rename(DpM=new_deaths_smoothed_per_million) %>% rename(VpH=people_fully_vaccinated_per_hundred)->k

scale.factor=max(k$VpH,na.rm=TRUE)/max(k$DpM,na.rm=TRUE)

ggplot(k,aes(x=as.Date(date)))+geom_line(aes(x=as.Date(date),y=DpM),color="red")+geom_point(aes(y=(VpH/scale.factor)), color="blue")+scale_y_continuous(name = expression("Deaths per million"),sec.axis = sec_axis(~ . *scale.factor, name = "Vaccination per hundred"))+labs(x="Date", title = paste(x))+theme_classic()-> f
  
print(f)
  
}


#matlines(x=vt,y=covid19china,type="l",col =ncol(covid19esa)+1 )

## Add legend
#legend(x="left",legend=c(names(covid19)), pch = 12, col = 1:ncol(covid19), bty = "n",horiz = 0,cex = .7)

```

```{=tex}
\newpage
\blandscape
```
```{r COVID3, echo=FALSE,include=FALSE, out.width='\\textwidth', fig.align = "center",fig.height = 7, fig.width = 9 }
days_since<-100
totaldb<-list()
countrynames<-c()
for(j in seq(1,ncol(covid19))) {

 cntryseries<-covid19[covid19[[j]]>=days_since,j]

 if (length(cntryseries)<nrow(covid19) & length(cntryseries)>0){
     cntryseries[(length(cntryseries)+1):(nrow(covid19))]<-NA

     assign(paste0(names(covid19)[j]),cntryseries)
     
     totaldb<-cbind(totaldb,assign(paste0(names(covid19)[j]),cntryseries))
     countrynames<-c(countrynames,names(covid19)[j])
 } else if (length(cntryseries)==nrow(covid19))  {
   totaldb<-cbind(totaldb,assign(paste0(names(covid19)[j]),cntryseries))
   countrynames<-c(countrynames,names(covid19)[j])
 }
 totaldb<-as.data.frame(totaldb)

 names(totaldb)<-countrynames
}

 totaldb<-as.data.frame(sapply(totaldb, as.numeric))


options(scipen=999)

vtx<-seq(0,nrow(totaldb)-1)
mult.fig(ncol(totaldb),main=paste("ESA Country Trends"))
for (y in names(totaldb)){
  
  matplot(x = vtx, y = log((totaldb)), type = "l", xlab = paste("Days since", days_since, "case"), ylab = "Log scale", main = y,lwd = 2, lty = 3, bty = "l", col = "grey", xaxt='n', yaxt='n', cex.main=1.2, cex.lab=.9)
  
  

matlines(x=(vtx), y=log(totaldb[[paste(y)]]), col="red",lwd = 2, lty = 1, bty = "l",xaxt='n', ylim=c(0,max(log(totaldb[[paste(y)]]))))
  #par(cex.main=.5)
  axis(side =1, cex.axis=.5)
  
  }
##plot all countries


#matlines(x=vt,y=covid19china,type="l",col =ncol(covid19esa)+1 )

## Add legend
#legend(x="left",legend=c(names(covid19)), pch = 12, col = 2:ncol(covid19), bty = "n",horiz = 0,cex = .72)

```

\elandscape
